---
title: "Blog Post 2"
subtitle: "Progress Discussion and Dataset Investigating"
description:  |
  Data Link:
  https://www.kaggle.com/datasets/konradb/prison-population-in-the-us?select=populations_states.csv
author: "Group 10"
date: "2024-10-18"
categories: [Blog Post 2]
date-modified: "2024-10-18"
draft: FALSE
---

### In response to our first blog post, we shifted our focus to analyzing the US prison population. This dataset includes information on admission and release totals, along with the total broken down by race, the date of this data collection, and the states from which the data was gathered. The data comes from the article titled “COVID-19 Amplified Racial Disparities in the US Criminal Legal System,” collected by a team of researchers. This data was collected to examine how the COVID-19 pandemic affected the racial composition of US prisons. Despite an overall decrease in the prison population, the proportion of incarcerated Black and Latino prisoners increased. So, the study aims to understand why this happened and how the pandemic made racial inequalities in the prison system worse. The data is available on Nature.com, a scientific journal that publishes peer-reviewed research and news.

### We firstly found the dataset on Kaggle. There is a link to github(https://github.com/jkbren/incarcerated-populations-data/), which is also the original source of this dataset. Also, there is a nature article discussing about the dataset (link https://www.nature.com/articles/s41586-023-05980-2)).

### The question asked most about this dataset is "What are the long-term trends in admissions and releases across different states, and how do these trends correlate with changes in policy". This is also the concern of policymakers.

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
```

```{r}
df = read.csv("admissions_releases_states.csv")
str(df)
```
### Next, we read the CSV file admissions_releases_states.csv into a dataframe df.
### We applied str() to display the structure of the dataset, including the names and types of the columns.

```{r}
df$date = as.Date(df$date, format="%Y-%m-%d")
```
### We converted the date column from a string format to a proper Date format using as.Date().
### This allows for easier date-based operations in the future.

```{r}
data = df %>%
  rename(
    Date = date,
    Total_Admissions = admissions_total,
    White_Admissions = admissions_white,
    Black_Admissions = admissions_black,
    Hispanic_Admissions = admissions_hispanic,
    AmericanIndian_Admissions = admissions_amerind,
    Asian_Admissions = admissions_asian,
    Other_Admissions = admissions_other,
    Total_Releases = releases_total,
    White_Releases = releases_white,
    Black_Releases = releases_black,
    Hispanic_Releases = releases_hispanic,
    AmericanIndian_Releases = releases_amerind,
    Asian_Releases = releases_asian,
    Other_Releases = releases_other,
    State = state
  )
```
### We applied rename() to change the column names to more representative and understandable terms, making the data clearer to work with.


```{r}
sapply(data, function(x) sum(is.na(x)))
```
### We applied sapply() to check for any missing values in the dataset. This function returns the number of NA (missing) values for each column.



```{r}
data = data %>% distinct()
```
### We removed any duplicate rows in the dataset by applying the distinct() function, ensuring each row in data is unique.


```{r}
summary(data)
```
### Finally, we did a statistical summary of the cleaned dataset using the summary() function, which provides basic statistics like the minimum, maximum, and mean values for numerical columns.


```{r}
saveRDS(data, "dataset/cleaned_dataset.rds")
```
### We saved the cleaned dataset as an .rds file using saveRDS().
### This will allow for quick loading of the cleaned dataset in future sessions using read_rds().


```{r}
admissions_totals = data %>%
  summarise(
    Total_White_Admissions = sum(White_Admissions, na.rm = TRUE),
    Total_Black_Admissions = sum(Black_Admissions, na.rm = TRUE),
    Total_Hispanic_Admissions = sum(Hispanic_Admissions, na.rm = TRUE),
    Total_AmericanIndian_Admissions = sum(AmericanIndian_Admissions, na.rm = TRUE),
    Total_Asian_Admissions = sum(Asian_Admissions, na.rm = TRUE),
    Total_Other_Admissions = sum(Other_Admissions, na.rm = TRUE)
  )

releases_totals = data %>%
  summarise(
    Total_White_Releases = sum(White_Releases, na.rm = TRUE),
    Total_Black_Releases = sum(Black_Releases, na.rm = TRUE),
    Total_Hispanic_Releases = sum(Hispanic_Releases, na.rm = TRUE),
    Total_AmericanIndian_Releases = sum(AmericanIndian_Releases, na.rm = TRUE),
    Total_Asian_Releases = sum(Asian_Releases, na.rm = TRUE),
    Total_Other_Releases = sum(Other_Releases, na.rm = TRUE)
  )
```

### Here, we calculated the total admissions and releases for each racial group by summing the values across the entire dataset. This ensures we can visualize the total counts of admissions and releases for different racial groups over the entire time range.

```{r}
admissions_melt = tidyr::gather(admissions_totals, key = "Race", value = "Admissions")
releases_melt = tidyr::gather(releases_totals, key = "Race", value = "Releases")


admissions_melt$Race = c("White", "Black", "Hispanic", "American Indian", "Asian", "Other")
releases_melt$Race = c("White", "Black", "Hispanic", "American Indian", "Asian", "Other")
```
### Here, we reshaped the summarized data using tidyr::gather() to convert it into a long format, which is suitable for plotting with ggplot2. Additionally, we ensured the race names were labeled correctly for clear and consistent x-axis labels in the plots.

```{r fig.height = 6, fig.width= 10}
ggplot(admissions_melt, aes(x = reorder(Race, -Admissions), y = Admissions, fill = Race)) +
  geom_bar(stat = "identity") +
  labs(title = "Total Prison Admissions by Race for the Entire Time Range", x = "Race", y = "Total Admissions") +
  theme_minimal() +
  theme(
    axis.title.x = element_text(margin = margin(t = 10)), 
    axis.title.y = element_text(margin = margin(r = 10))
  ) +
  scale_fill_manual(values = c(
    "White" = "#1f77b4",    
    "Black" = "#ff7f0e",    
    "Hispanic" = "#2ca02c", 
    "American Indian" = "#d62728", 
    "Asian" = "#9467bd",    
    "Other" = "#8c564b"   
  ))
```
### In this plot, we visualized the total prison admissions by race across the entire time range from January 1, 2000 to September 1, 2022. Each bar represents a different racial group, and we used custom colors for the bars to distinguish them. The bars are sorted in descending order based on the total admissions for each race. The bar plot for total prison admissions by race shows that White individuals account for the highest number of admissions (865,467), followed by Black individuals (368,346), and Hispanic individuals (277,007). American Indian admissions (36,958) are notably lower, and Asian (7,396) and Other (18,627) groups represent the smallest proportions. These disparities highlight racial differences in prison admissions, and future analyses will explore how these patterns change over time, particularly focusing on significant periods such as the COVID-19 pandemic.



```{r  fig.height = 6, fig.width= 10}
ggplot(releases_melt, aes(x = reorder(Race, -Releases), y = Releases, fill = Race)) +
  geom_bar(stat = "identity") +
  labs(title = "Total Prison Releases by Race for the Entire Time Range", x = "Race", y = "Total Releases") +
  theme_minimal() +
  theme(
    axis.title.x = element_text(margin = margin(t = 10)), 
    axis.title.y = element_text(margin = margin(r = 10))
  ) +
  scale_fill_manual(values = c(
    "White" = "#1f77b4",  
    "Black" = "#ff7f0e",  
    "Hispanic" = "#2ca02c", 
    "American Indian" = "#d62728",
    "Asian" = "#9467bd",  
    "Other" = "#8c564b" 
  ))
```
### Similarly, this plot shows the total prison releases by race for the entire time range from January 1, 2000 to September 1, 2022. Like the admissions plot, the bars are color-coded and sorted in descending order of total releases, giving a clear overview of the distribution of releases among different racial groups. The bar plot for total prison releases shows similar trends, with White individuals having the highest release count (899298), followed by Black (407339) and Hispanic (307995) individuals. Again, American Indian (37153), Asian (7540), and Other (22,563) groups have smaller totals. Moving forward, we will conduct a time-dependent analysis to examine trends over various time periods, with special attention to the effects of COVID-19 on prison admissions and releases during the pandemic.

### Transparency and representation are key principles in equitable data practices, both of which are relevant to our analysis of the prison admissions and releases dataset. We will ensure transparency by clearly documenting the data source, collection methods, and any potential limitations, such as missing data or sample bias, to prevent misuse or misinterpretation. Additionally, we will focus on representation by checking the distribution of different racial groups in the dataset to ensure no group is underrepresented, and we will acknowledge any imbalances that may affect our results. This approach helps promote fairness and accuracy in our analysis.